---
title: "HPOExplorer" 
author: "<h4>Authors: <i>Robert Gordon-Smith, Brian M. Schilder, Nathan G .Skene</i></h4>" 
date: "<h4>Most recent update: <i>`r format( Sys.Date(), '%b-%d-%Y')`</i></h4>"
output:
  github_document 
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

`r badger::badge_github_version(color="black")`
`r badger::badge_github_actions()`
`r badger::badge_codecov()`
`r badger::badge_last_commit()`
`r badger::badge_license()`

## Intro

This package contains useful functions for working with the [Human Phenotype Ontology (HPO)](https://hpo.jax.org/app/). It allows you to create interactive phenotype network plots, as well as many other useful functions.

## Installation

Within R: 

```{r, eval=FALSE}
if(!"remotes" %in% rownames(install.packages())){install.packages("remotes")}

remotes::install_github("neurogenomics/HPOExplorer")
```

## Documentation 

Visit the package site for a tutorial and docs for functions.  
[`HPOExplorer` site](https://neurogenomics.github.io/HPOExplorer/)

# Session Info 

<details> 

```{r Session Info}
utils::sessionInfo()
```

</details>  
<hr>
