---
title: "" 
author: "`r rworkflows::use_badges()`" 
date: "<h4>Most recent update: <i>`r format( Sys.Date(), '%b-%d-%Y')`</i></h4>"
output:
  github_document 
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pkg <- read.dcf(file="DESCRIPTION",
                fields = "Package")[[1]]
```
 
## Intro

`HPOExplorer` contains useful functions for working with the
[Human Phenotype Ontology (HPO)](https://hpo.jax.org/app/). 
It allows you to create interactive phenotype network plots,
as well as many other useful functions.

## Installation

Within R: 

```{r, eval=FALSE}
if (!require("remotes")) install.packages("remotes")
if (!require("HPOExplorer")) remotes::install_github("neurogenomics/HPOExplorer")

library(HPOExplorer)
```

## [Documentation website](https://neurogenomics.github.io/HPOExplorer/)

### [Getting started](https://neurogenomics.github.io/HPOExplorer/articles/HPOExplorer.html)

A quick tutorial on how to get started with `HPOExplorer`.

### [Docker](https://neurogenomics.github.io/HPOExplorer/articles/docker.html) 

`HPOExplorer` is also available via [DockerHub](https://hub.docker.com/repository/docker/neurogenomicslab/hpoexplorer). 
Click
[here](https://neurogenomics.github.io/HPOExplorer/articles/docker.html)
for instructions on how to create a Docker or Singularity
container with `HPOExplorer` and Rstudio pre-installed. 

# Citation

If you use ``r pkg``, please cite: 

<!-- Modify this by editing the file: inst/CITATION  -->
> `r utils::citation(pkg)$textVersion`

# Session Info 

<details> 

```{r Session Info}
utils::sessionInfo()
```

</details>  
<hr>
